The contact value approximation for the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-06-11 papers/pair-correlation/paper.tex
work on pairs paper, add todos Tested on bennet. Took 0:18.96.

@@

\section{Previous theoretical approaches}

The classic (and earliest) approach for computing the pair
distribution function is to use Percus' trick of treating one sphere
as an external field, and to use the resultant equilibrium density to
find the pair distribution function~\cite{hansen2006theory}.  This
elegant approach lends itself to DFT, and can be used to compute and
plot the pair distribution function, but requires a full free-energy
minimization \emph{for each position} $\rr_1$ in
$g^{(2)}(\rr_1,\rr_2)$, and hence would be prohibitively expensive as
a tool in constructing a free energy functional.

The canonical inhomogeneous configuration for the hard-sphere fluid is
the system consisting of a hard sphere at a hard wall.  In 1986,
Plischke and Henderson solved the pair distribution function of this
system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall by solving the Ornstein-Zernike equation using the direct
correlation function found from a functional derivative of a classical
DFT free energy functional~\cite{gotzelmann1996structure}.
Unfortunately, solving the inhomogeneous Ornstein-Zernike equation is
computationally challenging, although more efficient approximate
algorithms have been developed~\cite{paul2003variational}.

@@

\section{Previous approximations to the pair distribution function}
There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  The approximations differ in what density
is used when evaluating the radial distribution function $g(r,\eta)$
or in how radial distributions evaluated at different densities are
combined to yield an effective function
\cite{toxvaerd1973statistical}.

Early approximations to the pair distribution function used the
density at one or two positions to determine the pair distribution
function.  There are three common approaches:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12};n\left(\frac{\rr_1+\rr_2}{2}\right)\right)
     & \text{midpoint} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12};\frac{n(\rr_1)+n(\rr_2)}{2}\right)
     & \text{mean density} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  \frac{g(r_{12},n(\rr_1))+g(r_{12},n(\rr_2))}{2}
     & \text{mean function}
\end{align}
These approaches have been successfully and widely used in treating
the surface tension of simple fluids~\cite{pressing2003surface,
  salter2008statistical, bongiorno1975modified,
  toxvaerd1976hydrostatic, kalos1977structure, carey2008gradient,
  osborn1980monotonic, mccoy1981comparison, barrett2006some}.  The
mean density approximation has also been quoted (as a goal) by recent
papers that proceed to make further
approximations~\cite{gloor2007prediction, gross2009density}.  While
these approximations fail dramatically when applied to strongly
inhomogeneous systems such as a dense fluid at a solid surface.  Such
systems exhibit a strongly oscillatory density distribution, with
density peaks that can have local packing fractions greater than
unity, which cannot occur in the bulk reference system that defines
$g(r, n)$.  The above papers restrict themselves to the liquid-vapor
interface, which does not exhibit this pathology.

@@

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the hard-sphere radial distribution function, with
    our separable fit.  The solid lines show the true radial
    distribution function $g(r)$, and the dashed lines show our
    separable fit to the radial distribution functions
    $g_S(r)$.}\label{fig:radial-distribution}
\end{figure}

@@

We construct $g_S(r, g_\sigma)$ to satisfy several constraints.
Naturally, we match the value at contact:
\begin{equation}
  g_S(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function, $h(r) = g(r) - 1$:
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $\chi_T$ is the isothermal compressibility of the hard-sphere fluid.
We have found that the slope of the radial distribution function at
contact is given to a very good approximation by:
\begin{align}
  g'(\sigma) \approx g_\sigma (1 - g_\sigma)
\end{align}
and we apply this as a final constraint.\par
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 in steps of 0.05.  Our fitted radial
distribution function is given by
\begin{multline}
  g_S(r,g_\sigma) = 1 + (g_\sigma-1) e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + (g_\sigma -1)(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} + (1-g_\sigma) \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    + (1-g_\sigma)\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
Finally, we can convert between the packing fraction $\eta$ and the
radial distribution function at contact $g_\sigma$ using the
Carnahan-Starling approximation for $g_\sigma$:
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{align}
  g_\sigma &= \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}\text{, and its inverse} \\
  \eta &= 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}.
\end{align}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the constraint on the integral of $h(r)$ from
Eq.~\ref{eq:total-constraint} results in canceling errors at large
distances.  The maximum error in $g_S(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr R$.

@@

>>>>

\fixme{replace approximation with approach}

@@

\fixme{Somewhere above introduce the question of whether g2 is
  expressed in terms of V or n}

\section{Previous theoretical approaches $g^{(2)}(\rr_1,\rr_2)[V(\rr)]$}

\fixme{introduce this as methods given V}

The classic (and earliest) approach for computing the pair
distribution function is to use Percus' trick of treating one sphere
as an external field, and to use the resultant equilibrium density to
find the pair distribution function~\cite{hansen2006theory}.  This
elegant approach lends itself to computation with DFT, and can be used to compute and
plot the pair distribution function, but requires a full free-energy
minimization \emph{for each position} $\rr_1$ in
$g^{(2)}(\rr_1,\rr_2)$, and hence would be prohibitively expensive as
a tool in constructing a free energy functional.

The canonical inhomogeneous configuration for the hard-sphere fluid is
the system consisting of a hard sphere at a hard wall.  In 1986,
Plischke and Henderson solved the pair distribution function of this
system using integral equation theory under the Percus-Yevick
approximation~\cite{plischke1986pair}.  \fixme{Move the following to
  the next section} Ten years later,
G{\"o}tzelmann \emph{et al.} found the pair distribution function near
a hard wall by solving the Ornstein-Zernike equation using the direct
correlation function found from a functional derivative of a classical
DFT free energy functional~\cite{gotzelmann1996structure}.
Unfortunately, solving the inhomogeneous Ornstein-Zernike equation
remains computationally challenging, although more efficient
approximate algorithms have been developed~\cite{paul2003variational}.
This approach, while appealing, remains unsuitable for use in the
construction of a classical density functional due to its significant
computational cost.

@@

\section{Previous approximations to the pair distribution function
  $g^{(2)}(\rr_1,\rr_2)[n(\rr)]$}

\fixme{introduce this as methods given n}

There are several analytic approximations for the inhomogeneous pair
distribution function, which extend the radial distribution function
to inhomogeneous scenarios.  The approximations differ in what density
is used when evaluating the radial distribution function $g(r,\eta)$
or in how radial distributions evaluated at different densities are
combined to yield an effective function
\cite{toxvaerd1973statistical}.

Early approximations to the pair distribution function used the
density at one or two positions to determine the pair distribution
function.  There are three common approaches:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}, n\left(\frac{\rr_1+\rr_2}{2}\right)\right)
     & \text{midpoint} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}, \frac{n(\rr_1)+n(\rr_2)}{2}\right)
     & \text{mean density} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  \frac{g(r_{12},n(\rr_1))+g(r_{12},n(\rr_2))}{2}
     & \text{mean function}
\end{align}
These approaches have been successfully and widely used in treating
the surface tension of simple fluids~\cite{pressing2003surface,
  salter2008statistical, bongiorno1975modified,
  toxvaerd1976hydrostatic, kalos1977structure, carey2008gradient,
  osborn1980monotonic, mccoy1981comparison, barrett2006some}.  The
mean density approximation has also been quoted (as a goal) by recent
papers that proceed to make further
approximations~\cite{gloor2007prediction, gross2009density}.  While
these approximations fail dramatically when applied to strongly
inhomogeneous systems such as a dense fluid at a solid surface.  Such
systems exhibit a strongly oscillatory density distribution, with
density peaks that can have local packing fractions greater than
unity, which cannot occur in the bulk reference system that defines
$g(r, n)$.  The above papers restrict themselves to the liquid-vapor
interface, which does not exhibit this pathology.

@@

\davidsays{Please edit!}\fixme{Note that the HS free energy is used
  for two purposes when applying Eq.~\ref{eq:gsigma-exact} : it is
  minimized (together with an external potential) to find the density
  $n(\rr)$.  Then its functional derivative is used to find
  $g_\sigma(\rr)$.  In application of perturbation theory, the HS
  functional would only be used alone for the latter purpose, as the
  density would be determined by the total free energy, which
  incorporates both the HS free energy and a perturabative
  contribution.}

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
  %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
  \caption{Plot of the hard-sphere radial distribution function, with
    our separable fit.  The solid lines show the true radial
    distribution function $g(r)$, as found from Monte Carlo
    simulation, and the dashed lines show our separable fit to the
    radial distribution functions
    $g_S(r)$.}\label{fig:radial-distribution}
\end{figure}

@@

We construct $g_S(r, g_\sigma)$ to satisfy several constraints.
Naturally, we match the value at contact:
\begin{equation}
  g_S(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function, $h(r) = g(r) - 1$:
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $\chi_T$ is the isothermal compressibility of the hard-sphere fluid.
We have found that the slope of the radial distribution function at
contact is given to a very good approximation by:
\begin{align}
  g'(\sigma) \approx g_\sigma (1 - g_\sigma)
\end{align}
and we apply this as a final constraint.\par
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 in steps of 0.05.  Our fitted radial
distribution function is given by
\begin{multline}
  g_S(r,g_\sigma) = 1 + (g_\sigma-1) e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + (g_\sigma -1)(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} + (1-g_\sigma) \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    + (1-g_\sigma)\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
where the dimensionless isothermal compressibility $\chi$ is obtained
from the Carnahan-Starling equation of state.
Finally, we can convert between the packing fraction $\eta$ and the
radial distribution function at contact $g_\sigma$ using the
Carnahan-Starling approximation for $g_\sigma$:
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{align}
  g_\sigma &= \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}\text{, and its inverse} \\
  \eta &= 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}.
\end{align}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the constraint on the integral of $h(r)$ from
Eq.~\ref{eq:total-constraint} results in canceling errors at large
distances.  The maximum error in $g_S(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr R$.

@@

\subsection{Efficiency}

\fixme{Explain why the separable version is efficient.}
